---
title: "Kriging Computing "
author: "Khumbelo"
date: "2023-09-19"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---
library(geoR)
library(fields)
library(sp)

```{r}
points(sic.100, borders=sic.borders,col="green")
points(sic.367, borders=sic.borders,col="red",add=TRUE)
```
```


```{r}

# Create a data frame with your data points
data <- data.frame(
  x = c(100, 367),
  y = c(100, 367)
)

# Calculate summary statistics
summary(data)
```




```{r}
plot.geodata(sic.100,bor=sic.borders)

```
```{r}
# Create a data frame with your data points
data <- data.frame(
  x = c(100, 367),
  y = c(100, 367)
)

# Calculate summary statistics
summary(data)

```


```{r}
vario.b<- variog(sic.100,option =c ("bin", "cloud", "smooth"),
bin.cloud=TRUE)
## variog: computing omnidirectional variogram
plot(vario.b)
```
```{r}
vario.c <- variog(sic.100, op="cloud")
## variog: computing omnidirectional variogram
plot(vario.c)
vario.c <- variog(sic.100, op="cloud")
## variog: computing omnidirectional variogram
bplot.xy(vario.c$u,vario.c$v, breaks=vario.b$u,col="grey80",
lwd=2,cex=0.1,outline=FALSE)


```
```{r}
plot(vario.c)

```


```{r}
vario.ex<- variog(sic.100, bin.cloud=TRUE)
## variog: computing omnidirectional variogram
plot(vario.ex)
```
```{r}
vario4<-variog4(sic.100)
plot(vario4,same=FALSE)
```




```{r}
vario.ex<- variog(sic.100,option="bin")

vario.sphe<-(variofit(vario.ex,cov.model= "spher",
ini.cov.pars=c(15000,200)))

par(mfrow=c(2,2), mar=c(3,3,1,1), mgp =c (2,1,0))
plot(vario.ex,main="Spherical")
lines.variomodel(cov.model="sphe",cov.pars=c(15000,100),
nug=0,max.dist=350)
plot(vario.ex,main="Exponential")
lines.variomodel(cov.model="exp",cov.pars=c(15000,100),
nug=0,max.dist=350)

```
```{r}
vario.ex<- variog(sic.100,option="bin")
vario.matern<-(variofit(vario.ex,cov.model= "spher",
ini.cov.pars=c(15000,200)))

par(mfrow=c(2,2), mar=c(3,3,1,1), mgp =c (2,1,0))
plot(vario.ex,main="Spherical")
lines.variomodel(cov.model="sphe",cov.pars=c(15000,100),
nug=0,max.dist=350)
plot(vario.ex,main="Exponential")
lines.variomodel(cov.model="exp",cov.pars=c(15000,100),
nug=0,max.dist=350)


```
```{r}
library(geoR)
kc1<- krige.conv(sic.100, loc = sic.100$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
kc2<- krige.conv(sic.100, loc = sic.367$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
plot(sic.100$data,kc1$predict,xlab="Observed",ylab="Estimated",
main="Control sample")
abline(a=0,b=1,col="red")
plot(sic.367$data,kc2$predict,,xlab="Observed",ylab="Estimated",
main="Control")
abline(a=0,b=1,col="red")

```

```{r}
# Function to calculate RMSE
calculate_rmse <- function(observed, predicted) {
  residuals <- observed - predicted
  rmse <- sqrt(mean(residuals^2))
  return(rmse)
}

# Calculate RMSE for kc1
rmse_kc1 <- calculate_rmse(sic.100$data, kc1$predict)
cat("RMSE for Control Sample:", rmse_kc1, "\n")

# Calculate RMSE for kc2
rmse_kc2 <- calculate_rmse(sic.367$data, kc2$predict)
cat("RMSE for Control:", rmse_kc2, "\n")

```



```{r}
# Compute the omnidirectional variogram
vario.ex <- variog(sic.100, option = "bin")

# Fit spherical model
vario.sphere <- variofit(vario.ex, cov.model = "sph", ini.cov.pars = c(15000, 200))

# Fit exponential model
vario.exp <- variofit(vario.ex, cov.model = "exp", ini.cov.pars = c(15000, 100))

# Fit Matérn model (with initial values for sigma^2 and phi)
vario.matern <- variofit(vario.ex, cov.model = "mat", ini.cov.pars = c(0.1, 15000))

# Fit Gaussian model (with initial values for sigma^2 and phi)
vario.gaussian <- variofit(vario.ex, cov.model = "gaussian", ini.cov.pars = c(0.1, 15000))

# Calculate RMSE for spherical model
rmse.sphere <- sqrt(mean((vario.ex$gamma - vario.sphere$gamma)^2))

# Calculate RMSE for exponential model
rmse.exp <- sqrt(mean((vario.ex$gamma - vario.exp$gamma)^2))

# Calculate RMSE for Matérn model
rmse.matern <- sqrt(mean((vario.ex$gamma - vario.matern$gamma)^2))

# Calculate RMSE for Gaussian model
rmse.gaussian <- sqrt(mean((vario.ex$gamma - vario.gaussian$gamma)^2))

# Plot variograms with fitted models and RMSE
par(mfrow = c(2, 2), mar = c(3, 3, 1, 1), mgp = c(2, 1, 0))
plot(vario.ex, main = "Spherical")
lines.variomodel(vario.sphere, max.dist = 350)
text(200, 0.1, paste("RMSE =", round(rmse.sphere, 2)))

plot(vario.ex, main = "Exponential")
lines.variomodel(vario.exp, max.dist = 350)
text(200, 0.1, paste("RMSE =", round(rmse.exp, 2)))

plot(vario.ex, main = "Matérn")
lines.variomodel(vario.matern, max.dist = 350)
text(200, 0.1, paste("RMSE =", round(rmse.matern, 2)))

plot(vario.ex, main = "Gaussian")
lines.variomodel(vario.gaussian, max.dist = 350)
text(200, 0.1, paste("RMSE =", round(rmse.gaussian, 2)))



```










```{r}
# Compute the omnidirectional variogram
vario.ex <- variog(sic.100, option = "bin")

# Fit spherical model without nugget effect
vario.sphere <- variofit(vario.ex, cov.model = "sph", ini.cov.pars = c(15000, 200))
rmse.sphere_without_nugget <- sqrt(mean((vario.ex$gamma - vario.sphere$gamma)^2))

# Fit spherical model with nugget effect
vario.sphere_nugget <- variofit(vario.ex, cov.model = "sph", nugget = TRUE, ini.cov.pars = c(15000, 200))
rmse.sphere_with_nugget <- sqrt(mean((vario.ex$gamma - vario.sphere_nugget$gamma)^2))

# Repeat the process for other models (exponential, Matérn, Gaussian)...

# Print RMSE values
cat("Spherical without Nugget RMSE:", round(rmse.sphere_without_nugget, 2), "\n")
cat("Spherical with Nugget RMSE:", round(rmse.sphere_with_nugget, 2), "\n")

# Repeat for other models...

```
```{r}
# Function to calculate RMSE, handling NaN values
calculate_rmse <- function(observed, predicted) {
  squared_diff <- (observed - predicted)^2
  mean_squared_diff <- mean(squared_diff, na.rm = TRUE)
  if (!is.finite(mean_squared_diff)) {
    return(NA)  # Return NA if the result is NaN or infinite
  }
  return(sqrt(mean_squared_diff))
}

# Compute the omnidirectional variogram
vario.ex <- variog(sic.100, option = "bin")

# Fit spherical model without nugget effect
vario.sphere <- variofit(vario.ex, cov.model = "sph", ini.cov.pars = c(15000, 200))
rmse.sphere_without_nugget <- calculate_rmse(vario.ex$gamma, vario.sphere$gamma)

# Fit spherical model with nugget effect
vario.sphere_nugget <- variofit(vario.ex, cov.model = "sph", nugget = TRUE, ini.cov.pars = c(15000, 200))
rmse.sphere_with_nugget <- calculate_rmse(vario.ex$gamma, vario.sphere_nugget$gamma)

# Print RMSE values
cat("Spherical without Nugget RMSE:", round(rmse.sphere_without_nugget, 2), "\n")
cat("Spherical with Nugget RMSE:", round(rmse.sphere_with_nugget, 2), "\n")

```





```{r}
library(geoR)
pred.grid <- expand.grid(seq(0,350, l=51),seq (0,220, l=51))
rgb.palette <- colorRampPalette(c("blue", "lightblue",
"orange", "red"),space = "rgb")
kc<- krige.conv(sic.100, loc = pred.grid,
krige=krige.control(cov.model="spherical",cov.pars=c(15000,50)))
image(kc, loc = pred.grid,col =rgb.palette(20) ,xlab="Coord X",
ylab="Coord Y",borders=sic.borders,main="Estimation")
image(kc, krige.var,loc = pred.grid,col=rgb.palette(20),
xlab="Coord X",ylab="Coord Y",borders=sic.borders,
main="Kriging variance")
```





```{r}
library(geoR)

# Assuming sic.100 and sic.borders are already defined
pred.grid <- expand.grid(seq(0, 350, length = 51), seq(0, 220, length = 51))
rgb.palette <- colorRampPalette(c("blue", "lightblue", "orange", "red"), space = "rgb")

# Perform kriging
kc <- krige.conv(sic.100, loc = pred.grid, krige = krige.control(cov.model = "spherical", cov.pars = c(15000, 50)))

# Extract kriging variance
krige.var <- kc$krige.var

# Plot kriging estimation
image(kc, loc = pred.grid, col = rgb.palette(20), xlab = "Coord X", ylab = "Coord Y", borders = sic.borders, main = "Estimation")

# Plot kriging variance
image(kc, krige.var, loc = pred.grid, col = rgb.palette(20), xlab = "Coord X", ylab = "Coord Y", borders = sic.borders, main = "Kriging Variance")
```




```{r}
# Assuming sic.100 and sic.borders are already defined
pred.grid <- expand.grid(seq(0, 350, length = 51), seq(0, 220, length = 51))
rgb.palette <- colorRampPalette(c("blue", "lightblue", "orange", "red"), space = "rgb")

# Perform kriging
kc <- krige.conv(sic.100, loc = pred.grid, krige = krige.control(cov.model = "spherical", cov.pars = c(15000, 50)))

# Extract kriging variance
krige.var <- kc$krige.var

# Plot kriging estimation
image(kc, loc = pred.grid, col = rgb.palette(20), xlab = "Coord X", ylab = "Coord Y", borders = sic.borders, main = "Estimation")

# Add legend for kriging estimation
legend("topright", legend = "Estimation", fill = rgb.palette(20), title = "Legend")

# Plot kriging variance
image(kc, krige.var, loc = pred.grid, col = rgb.palette(20), xlab = "Coord X", ylab = "Coord Y", borders = sic.borders, main = "Kriging Variance")

# Add legend for kriging variance
legend("topright", legend = "Kriging Variance", fill = rgb.palette(20), title = "Legend")

# Custom legend with color labels
legend("bottomright", legend = c("Low", "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", "High"),
       fill = rgb.palette(20), bty = "n", title = "Color Legend")




```


```{r}
```
```{r}
library(geoR)
kc1<- krige.conv(sic.100, loc = sic.100$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
kc2<- krige.conv(sic.100, loc = sic.367$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
plot(sic.100$data,kc1$predict,xlab="Observed",ylab="Estimated",
main="Control sample")
abline(a=0,b=1,col="red")
plot(sic.367$data,kc2$predict,,xlab="Observed",ylab="Estimated",
main="Control")
abline(a=0,b=1,col="red")

```
```{r}
# Assuming kc1 and kc2 are defined as described in your code

# Calculate residuals for kc1
residuals_kc1 <- sic.100$data - kc1$predict

# Calculate RMSE for kc1
rmse_kc1 <- sqrt(mean(residuals_kc1^2))

# Calculate residuals for kc2
residuals_kc2 <- sic.367$data - kc2$predict

# Calculate RMSE for kc2
rmse_kc2 <- sqrt(mean(residuals_kc2^2))

# Print RMSE values
cat("RMSE for Control sample:", rmse_kc1, "\n")
cat("RMSE for Control:", rmse_kc2, "\n")

```



```{r}
library(geoR)
plot.geodata(sic.100,bor=sic.borders,lambda=0.5)

```


```{r}
# Create a data frame with your data points
data <- data.frame(
  x = c(100, 367),
  y = c(100, 367)
)

# Calculate summary statistics
summary(data)


```

```{r}
library(geoR)

# Assuming sic.100 and sic.borders are defined
plot.geodata(sic.100, bor = sic.borders, lambda = 0.5)

# Compute the omnidirectional variogram
vario.ex <- variog(sic.100, option = "bin")

# Fit spherical model
vario.sphere <- variofit(vario.ex, cov.model = "sph", ini.cov.pars = c(15000, 200))

# Fit exponential model
vario.exp <- variofit(vario.ex, cov.model = "exp", ini.cov.pars = c(15000, 100))

# Fit Matérn model (with initial values for sigma^2 and phi)
vario.matern <- variofit(vario.ex, cov.model = "mat", ini.cov.pars = c(0.1, 15000))

# Fit Gaussian model (with initial values for sigma^2 and phi)
vario.gaussian <- variofit(vario.ex, cov.model = "gaussian", ini.cov.pars = c(0.1, 15000))

# Plot variograms with fitted models
par(mfrow = c(2, 2), mar = c(3, 3, 1, 1), mgp = c(2, 1, 0))
plot(vario.ex, main = "Spherical")
lines.variomodel(vario.sphere, max.dist = 350)

plot(vario.ex, main = "Exponential")
lines.variomodel(vario.exp, max.dist = 350)

plot(vario.ex, main = "Matérn")
lines.variomodel(vario.matern, max.dist = 350)

plot(vario.ex, main = "Gaussian")
lines.variomodel(vario.gaussian, max.dist = 350)


```


```{r}
# Assuming sic.100 and sic.borders are defined
plot.geodata(sic.100, bor = sic.borders, lambda = 0.5)

# Compute the omnidirectional variogram
vario.ex <- variog(sic.100, option = "bin")

# Fit spherical model
vario.sphere <- variofit(vario.ex, cov.model = "sph", ini.cov.pars = c(15000, 200))

# Plot variogram with fitted spherical model
par(mfrow = c(1, 1), mar = c(3, 3, 1, 1), mgp = c(2, 1, 0))
plot(vario.ex, main = "Spherical")
lines.variomodel(vario.sphere, max.dist = 350)

```

```{r}
# Compute variogram using the first approach
vario.ex <- variog(sic.100, option = "bin")

# Fit exponential variogram model using the first approach
vario.exp <- variofit(vario.ex, cov.model = "exp", 
                      ini.cov.pars = c(15000, 100))

# Plot exponential variogram model using the first approach
plot(vario.ex, main = "Exponential (Approach 1)")
lines.variomodel(cov.model = "exp", cov.pars = c(15000, 100),
                 nug = 0, max.dist = 350)

# Model parameters for the second approach
modNugget <- 0        # Intercept-like term
modRange <- 100000    # Distance of the plateau
modSill <- 0.75       # Semivariance at the range

# Create variogram model using the second approach
ppt.vgm1 <- vgm(psill = modSill, 
                "Exp",        # Exponential model
                range = modRange, 
                nugget = modNugget)

# Fit variogram model to observed variogram using the second approach
ppt.vgm2 <- fit.variogram(ppt.var, ppt.vgm1)

# Plot observed variogram and fitted exponential variogram model using the second approach
plot(ppt.var, pptE.vgm2, 
     main = "Swiss Precip. Exponential Variogram Model (Approach 2)")

```




```{r}
library(geoR)
kct<- krige.conv(sic.100, loc = pred.grid,
krige=krige.control(cov.model="matern",cov.pars=c(105, 36),
kappa=1,nugget=6.9,lambda=0.5))
pred.grid <- expand.grid(seq(0,350, l=51),seq (0,220, l=51))
rgb.palette <- colorRampPalette(c("blue", "lightblue",
"orange", "red"),space = "rgb")
image(kct, loc = pred.grid,col =rgb.palette(20) , xlab="Coord X",
ylab="Coord Y",borders=sic.borders,main="Estimation")
image(kct, krige.var,loc = pred.grid,col =rgb.palette(20) ,
xlab="Coord X",ylab="Coord Y",borders=sic.borders,
main="Kriging variance")
```






```{r}
library(geoR)
pred.grid <- expand.grid(seq(0, 350, length.out = 51), seq(0, 220, length.out = 51))
rgb.palette <- colorRampPalette(c("blue", "lightblue", "orange", "red"), space = "rgb")
kc <- krige.conv(sic.100, loc = pred.grid,
                 krige = krige.control(cov.model = "spherical", cov.pars = c(15000, 50)))
image(kc, loc = pred.grid, col = rgb.palette(20), xlab = "Coord X",
      ylab = "Coord Y", borders = sic.borders, main = "Estimation")
krige.var <- kc$krige.var
image(kc, krige.var, loc = pred.grid, col = rgb.palette(20),
      xlab = "Coord X", ylab = "Coord Y", borders = sic.borders,
      main = "Kriging Variance")
```


```{r}
#   estimated and observed values}
library(geoR)
kc1<- krige.conv(sic.100, loc = sic.100$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
kc2<- krige.conv(sic.100, loc = sic.367$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
plot(sic.100$data,kc1$predict,xlab="Observed",ylab="Estimated",
main="Control sample")
abline(a=0,b=1,col="red")
plot(sic.367$data,kc2$predict,,xlab="Observed",ylab="Estimated",
main="Control")
abline(a=0,b=1,col="red")

```

```{r}
# Assuming kc1 and kc2 are defined as described in your code

# Calculate RMSE for kc1
rmse_kc1 <- sqrt(mean((sic.100$data - kc1$predict)^2))

# Calculate RMSE for kc2
rmse_kc2 <- sqrt(mean((sic.367$data - kc2$predict)^2))

# Print RMSE values
cat("RMSE for Con sample:", rmse_kc1, "\n")
cat("RMSE for Control Kriging:", rmse_kc2, "\n")

```

```{r}
kc1 <- krige.conv(sic.100, loc = sic.100$coords, krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
kc2 <- krige.conv(sic.100, loc = sic.367$coords, krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))

```


```{r}

library(geoR)
kc1<- krige.conv(sic.100, loc = sic.100$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
kc2<- krige.conv(sic.100, loc = sic.367$coords,
krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
plot(sic.100$data,kc1$predict,xlab="Observed",ylab="Estimated",
main="Control sample")
abline(a=0,b=1,col="red")
plot(sic.367$data,kc2$predict,,xlab="Observed",ylab="Estimated",
main="Control")
abline(a=0,b=1,col="red")



```

```{r}
# Assuming kc1 and kc2 are your kriging predictions
# Assuming sic.100 and sic.367 are your datasets
# Assuming kc1 and kc2 are your kriging predictions
# Assuming sic.100 and sic.367 are your datasets

# Function to calculate RMSE
calculate_rmse <- function(observed, predicted) {
  residuals <- observed - predicted
  rmse <- sqrt(mean(residuals^2))
  return(rmse)
}

# Calculate RMSE for kc1
rmse_kc1 <- calculate_rmse(sic.100$data, kc1$predict)
cat("RMSE for Control Sample:", rmse_kc1, "\n")

# Calculate RMSE for kc2
rmse_kc2 <- calculate_rmse(sic.367$data, kc2$predict)
cat("RMSE for Control:", rmse_kc2, "\n")



```


